# Multivariate General Linear Models

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>", echo = TRUE, message=FALSE, note=FALSE, warning=FALSE, cache=TRUE
)
```

Years ago I was chit-chatting with my dad. I don't remember how it came up, but we were talking about the death penalty. I am vehemently opposed to the death penalty. My dad, on the other hand, is in favor. As we debated the pros and cons, I mentioned  certain races were disproportionately affected. 

"Hogwash," he said. "I've seen the statistics. They say that black men are no more likely than white men to receive the death penalty."

"True," I said. I was familiar with the statistic he was quoting. 

But that's not the whole story. If a black man kills another black person, he is just as likely to get the death penalty as a white man who kills a white person. However, if a black man kills a *white* person, they are *much* more likely to receive the death penalty than if a white man kills a black person. 

That there is unfair, you see. (Not to mention the fact that it's government-sponsored killing, and all that). 

My dad's response? 

"Well son," he said, "That just goes to show you can use statistics to prove anything."

That pissed me off, I must admit. It was a dismissal of my argument--a protestation that said my statistics were no better than his. But, as we'll learn, simple bivariate effects (in this case, race and death penalty) are always less complete than multivariate effects (in this case, race of perpetrator *and* race of victim). 

## What is a multivariate relationship?

That there is a tricky question. And, you'll probably get different answers, depending on who you ask. Technically, multivariate means multiple variables. So, I suppose, any analysis that has more than one variable is a multivariate analysis. 

But, I've actually seen people refer to multivariate relationships in one of two ways:

1. *When one performs an analysis of multiple **dependent** variables.* When statisticians use the word multivariate, they tend to be referring to this sort of analysis, where multiple dependent variables are investigated. These situations call for using Multivariate Analysis of Variance (MANOVA), factor analysis, principle component analysis, structural equation modeling, and canonical correlations. I personally find many of these sorts of analyses archaic and quite useless. They are largely atheoretical, they confuse exploratory and confirmatory research, and they're hard to make sense of. That is *not* what this chapter is about. 

2. *When one performs an analysis of multiple **independent** variables.* Most everybody, with the exception of heritage breed statisticians, use the term multivariate to refer to situations where multiple independent variables are used. This is what *I* mean when I say multivariate in this textbook. Does that make me a mutt-breed statistician? Maybe. But, sometimes you have to sacrifice authenticity for practicality. 

So, yes, this chapter (and those that follow) are referring to situations where we have multiple predictor variables we are using to model a single outcome variable. Yay. 

## Reasons to use multivariate GLMs
 
The world is rarely simple. Rarely does one variable *alone* cause another variable. Rather, variables work together to produce causes. Or, one variable's effects are contingent upon another's. We use mutivariate GLMs to gain a more complete picture of what is influencing our outcome variable. 

But, that's a bit vague and stuff. Let me be a little more specific. *There are three reasons you would want to do a multivariate analysis*:

1. Study interaction effects. Sometimes variables "interact," which means their influence depends on other variable(s). For example, the annoyance I feel (outcome variable) about having department meetings (predictor variable #1) *depends* on whether there's food there (predictor variable #2); I'm happy to attend department meetings if they have baklava and pizza. Otherwise, bleeh. 

2. You want to *control* for things you're not interested in. For example, you might know depressed people tend to have crappy social lives, but you don't care about studying social functioning. You really care about depression's unique influence on, say, health. If that's the case, we want to "control" for social functioning. 

3. To improve predictions. Long story short, the more variables you have, the better your predictions. So, naturally, if you want to predict who will be the next world wood-chopping champion will be, you can add more predictors (e.g., bicep circumference, years of experience, beard length). 

Before we confabulate about each of these reasons, let's take a mini break to look at some pictures. Why not, eh? Pictures are pretty. 

Though, as to not bother my mythical editor, I'm going to embed these pretty pictures within the process of data analysis. 

## Visualizing Multivariate Relationships in Flexplot

In the previous section, I said there are three primary reasons for doing a multivariate analysis: to study interaction effects, to control for things, and to improve predictions. I'm going to temporarily put the third reason (prediction) aside and focus on interactions and conditioning. These two approaches have very different visualization strategies. But, before I can share those strategies, I need to show you some tools. 

### Encoding Additional Dimension Using Colors/Lines/Symbols or Panels

As a reminder, flexplot is essentially God-like in its ability to visualize data. For univariate data, remember all we have to do is write the variable name, followed by `~1`, then the dataset. For example...

```{r mvglmuniv}
require(flexplot)
data(avengers)
flexplot(speed~1, data=avengers)
flexplot(superpower~1, data=avengers)
```

For bivariate data, the pattern is similar, except we now put `outcome~predictor, data=dataset`. In the background, flexplot is going to figure out whether to plot a scatterplot or bee swarm plot:

```{r mvglmbivar}
flexplot(speed~superpower, data=avengers)
flexplot(speed~agility, data=avengers)
```

Yes, flexplot can read your mind. (Just make sure your computer is password protected. Flexplot hasn't yet figured out how to steal your secret recipes from your documents folder, but it's also pretty smart...so, yeah.)

For multivariate relationships, flexplot is even more powerful. In fact, when I originally conceptualized it, my whole purpose was to develop an easy interface for multivariate visualizations. So, that's where it truly shines. 

When trying to visualize interactions, it is important to see all the variables simultaneously. But how exactly do you visualize multivariate data? Alas, we humans can really only see in 3-4 dimensions (width, height, depth, and time). But that's not even true. We don't actually see depth. Instead, we rely on sharpness, size, and shading to mimic a third dimension. Even 3D movies aren't truly 3D--they just trick our vision into believing we're actually seeing in 3D. 

We *could* use the same tricks for visualization to see three-dimensional data. For example, 

```{r mvglm3d}
require(scatterplot3d)
data(avengers)
scatterplot3d(avengers$speed, avengers$agility, avengers$superpower)
```

(By the way, I wouldn't bother installing that package. We won't use it again.)

I just find these sorts of plots hard to understand. If I were able to view them as a hologram projection, I might find them useful. (Help us Obi Wan, you're our only hope!). But as a static picture, it's a bit tough to tell what's going on. And, 3D plots are limited to three variables; good luck visualizing four dimensions!

So what do we do instead? We can embed the third dimension in the same two-dimensional plot using a different mode of representation.

Lemme show you what I mean. Let's visualize the same data using flexplot:

```{r mvglmflex1}
flexplot(agility~speed + superpower, data=avengers, method="lm")
```

Notice how I've embedded a third variable (superpower) in the same plot as the two-dimensional data. From this, it's much easier to see the multivariate relationship: superheroes tend to decrease slightly in agility when speed increases, while non-superheroes tend to increase. 

Nice. 

Let's look at another example, but this time with two numeric variables:

```{r mvglmnumerictwo}
flexplot(agility~strength + speed, data=avengers, method="lm")
```

What did Flexplot do? Well, you done handed it a numeric variable and asked it to plot three variables at the same time. Since Flexplot refuses to resort to 3D plots, you gave it no other option. It decided to break speed into three equal segments (think of it as slow, average, and fast). Then it plotted the datapoints for the slow folks in red, the average folks in green, and the fast folks in blue. It plotted these levels as separate lines (dotted, dashed, and dot-dashed) and symbols (circles, squares, triangles). 

But now it's hard to see what's going on. Sure, we can see the lines alright, but the raw data are such a jumbled mess it's impossible to tell whether the regression lines actually fit the data.

Quite a conundrum, that. 

So, what to do? Well, let's investigate another technique called paneling. Paneling means that, rather than plotting the three levels of the strength variable as separate lines/colors/symbols, it plots each of the three levels in their own scatterplot (called a panel). To specify a panel, we use the code `y~x | z`. In other words, we replace the `+` with a `|`. That button's right above the return button. 

Let's take a look, see:

```{r mvglmpanel}
flexplot(agility~strength | speed, data=avengers)
```

Now we're talking! Now there's no overlaps and we can see that some of these relationships are actually quite curvilinear (particularly the strength/agility relationship for the slow folks).

Maybe we'll correct that curvilinearity with some quadratic lines:

```{r mvglmquadratic}
flexplot(agility~strength | speed, data=avengers, method="quadratic")
```

That's better.

But now we have a new problem; when the third variable was in the same panel, it was easy to make comparisons across the third variable. But now, it's much harder. For example, in the image above, which curved line is higher on agility? The average speed folks or the slow folks?

Hard to say. 

But, we can use a visual aid called a "ghost line." A ghost line repeats the line from one panel to another panel. Let's look at an example:

```{r mvglmghost}
flexplot(agility~strength | speed, data=avengers, 
         method="quadratic", 
         ghost.line="red")
```

Now we're getting somewhere! As you can see, the ghost line duplicates the relationship from the middle panel across to the other panels. Now it's very clear that the curve from the fast group (left-most plot) is lower, while the curve from the slow group (right-most plot) is higher in agility. 


```{block, type='rmdnote'}

*Why call it 'Ghost Line'?*

Ghost line is kind of a funny name, isn't it? How did I come up with that name? Well, back when I was a teenager, Nintendo 64 came it. That, my friends, was a brilliant videogaming system. One of its hallmark games was Mariokart. In Mariokart, you could choose which character to race with (e.g., Bowser, Mario, Luigi), then you would circle a track, trying to avoid shell-missiles. 

Anyhow, as I recall, whenever you were racing and set a speed record, the game would actually record your every move during the entire race. You then had the option to turn on the "ghost racer." This ghost racer was a semi-transparent image of the exact race that holds the record. The basic idea was that you could try to race against the best race ever recorded. 

Cool idea and cool concept. So much so, that I decided I would borrow the idea. Like in Mariokart, the ghost line doesn't actually reflect anything in that particular panel, but it is a representation of something outside the panel. 
```

Ghost lines, it turns out are essential for doing multivariate analyses, particularly when we start adding even *more* variables to the analysis. 

Let's look at another example:

```{r mvglmghost2}
flexplot(shots.taken~agility | speed + superpower, data=avengers, 
         method="quadratic", 
         ghost.line="red")
```

Notice how *nearly every single datapoint* for the superhero plots (the bottom panel) are below the fit of the ghost line. In other words, superheroes are taking far fewer shots than non-superheroes. We would not be able to see this without the ghost line. 

We'll see several examples in later chapters where the ghost line is essential to helping us figure out what's going on. 

By the way, you may have noticed that the ghost line generally picks the middle-most panel as the reference. But, you can tell Flexplot which panel to reference from. Suppose, for example, if we wanted to reference from the fast panel for superheroes. To do so, we would need to specify a list that tells flexplot the speed and superpower value to reference from:

```{r mvglmghostref}
flexplot(shots.taken~agility | speed + superpower, data=avengers, 
         method="quadratic", 
         ghost.line="red",
         ghost.reference = list(speed=4.5, superpower="yes"))
```

Here, we've chosen a value for speed that is in the range of the left-most panel (4.5 is within the 4.3-4.9 range) and told flexplot to choose "yes" as the superpower value. We could also specify slow non-superheroes:

```{r mvglmghostref2}
flexplot(shots.taken~agility | speed + superpower, data=avengers, 
         method="quadratic", 
         ghost.line="orange",
         ghost.reference = list(speed=5.2, superpower="no"))
```

Notice I also changed the color of the ghost line to orange. 

By the way, you can (realistically) visualize up to four predictor variables. For example:

```{r mvglmfourvars}
flexplot(shots.taken~agility + superpower | speed + strength, data=avengers, 
         method="lm", 
         ghost.line="black")
```

Now, Flexplot is representing the second variable as different lines/colors/symbols while paneling the remaining two variables (speed and strength). Also, the ghost line is referencing off of the middle-most panel for the superheroes.

Don't worry about interpreting this plot yet; I'll soon give you some cool tricks you can use to figure out what's going on here. The important take-home message for you is understanding how you can control the display of flexplot.

In summary, flexplot visualizes multivariate data by encoding additional variables as either a separate color/symbol/line, or in separate panels. To encode a variable as a color/line/symbol, use `y~x+z`. To encode as separate panels, use `y~x|z`. When we panel, it makes it hard to compare fits, so we can use "ghost lines" to repeat the pattern from one panel to another. 

Regardless of whether you encode additional variables as colors/symbols/lines or as separate panels, these sorts of plots are excellent for detecting interactions, which we'll discuss shortly. But, they can induce information overload. 

If you don't have interactions, these sorts of plots are needlessly complicated. As an alternative, we can use "added variable plots."

### Encoding Additional Dimensions Using Added Variable Plots

Earlier I said we used multivariate analyses for two primary purposes: (1) to identify interactions, and (2) to "control" for other variables. The previous section displayed the additional variables colors/symbols/lines or as separate panels. These plots are great for detecting interactions. If there are no interactions, we can simplify the visualization immensely by using added variable plots (or AVPs).

Before we talk about AVPs, though, let me offer some definitions of important terms. Let's say you know that one's social support structure affects health. You also know that depression affects health, but you want to see the unique effect of depression on health, above and beyond social structure. When doing this sort of analysis, we have three types of variables: the outcome variable (you already know what this is), the "control" variable, and the "interest" variable. 

**The control variable is the variable you're trying to control for.** Usually, we don't care about this variable. In our example, social support is the control variable. 

**The interest variable is the variable whose unique effect you're interested in studying.** In this example, our interest variable is depression, because that's the variables whose influence we're actually interested in studying. 
're trying to visualize the effect of depression on health, above and beyond s

You can have multiple control variables (e.g., what is the effect of depression on health, after controlling for a. social support, and b. exercise frequency). You could also have multiple interest variables, but this is less common. 

Let me summarize what AVPs do, before I go into more detail:

**AVPs fit a model between the control variable(s) and the outcome, then plot those residuals against the interest variable.**

Again, I'll go over this in detail and explain *why* we do this. But, before I lose your interest or before I lose you in the details, here's an example of an AVP in R:

```{r mvglmavp1}
added.plot(agility~speed + strength, data=avengers)
```

By default, AVP's control for the first variable(s) entered (speed), then plot the last variable entered (strength) on the x-axis. 

#### What Do AVPs Do and Why?

This will be easiest to explain with a simple example. Let's say we have the following dataset:

```{r mvglmtab, echo=FALSE}
set.seed(121211)
cor.mat = matrix(c(
  1, -.6, .9,
  -.6, 1, -.2,
  .9, -.2, 1
  ), nrow=3)
require(tidyverse)
d = data.frame(MASS::mvrnorm(n=10, mu=c(0,0,0), cor.mat))
names(d) = c("health", "depression", "support")
d = d %>% mutate(health = rescale(health, 20, 4) %>% floor_ceiling(10,30) %>% round,
             depression = rescale(depression, 12,5) %>% floor_ceiling(1, 30) %>% round, 
             support = rescale(support, 50, 15) %>% floor_ceiling(0) %>% round)

mod = lm(health~support, data=d)
knitr::kable(d)

```

Here, `support` is our control variable. Let's go ahead and look at a scatterplot of that model:

```{r mvglmavpfitmod, echo=FALSE}
flexplot(health~support, data=d, method="lm", se=F)
```

The regression line has the equation: $\text{health} = `r round(coef(mod)[1], digits=1)` + `r round(coef(mod)[2], digits=1)`\times \text{support}$. Let's go ahead and put each person's predicted score back in the table, as well as their residual:

```{r, echo=FALSE}
mod = lm(health~support, data=d)
d$prediction = predict(mod) %>% round(1)
d$residual = residuals(mod) %>% round(1)
knitr::kable(d)
```

Just for fun, go ahead and verify that the column labeled `prediction` is equal to `r coef(mod)[1]` + `r coef(mod)[2]`$\times$ support. Also, go ahead and verify that the column labeled `residual` is equal to health - prediction. 

Now, we have all we need to create an AVP. All we have to do is plot depression on the $X$ axis and the column labeled "residual" on the $Y$ axis:

```{r mvglmresiduals, echo=FALSE}
flexplot(residual~depression, data=d, method="lm", se=F)
```

What is this telling us? Remember what a residual means: *a residual tells us what remains in the outcome variable after removing the effect of whatever we modeled*. Here, we modeled the effect of social support on health. So, this residual represents health, after removing the effect of social support. 

And that, my friends, is exactly what we were looking for! Whenever we are trying to "control" or "condition" on some variable, the appropriate visual to use is an AVP. 

By the way, I didn't invent AVPs. (I wish I had!). But, I did modify them. I think most people would be confused looking at the plot above. So, flexplot makes two modifications: first, it relabels to axis (in this case, it would say `health | support`). This is because most people don't know what a residual is. Second, flexplot adds the mean of the outcome variable to every residual score. Why? Remember residuals always have a mean of zero. If the original variable had a mean of 15, then suddenly the plot showed it had a mean of zero, it would confuse people. So, I decided to add the mean of the outcome variable back in, to prevent confusion. 

So, in short, AVPs literally fit a linear model predicting the outcome from the control variable(s), compute the residuals, then plot the interest variable against the residuals. These plots visually show us the effect of the interest variable, after controlling for the control variable(s). 

#### Using Flexplot To Do AVPs

There's really two ways to use AVPs. One way is easy, but less flexible. The other is harder, but much more flexible. If you only have one interest variable and your control variables have simple relationships with the outcome, you can create AVPs like this:

```{r mvglmavpeasy}
added.plot(ptsd~agility + north_south, data=avengers)
```

The last variable entered (in this case, `north_south`) will be treated as the interest variable. All others will be control variables. Here's an example with multiple control variables:

```{r mvglmavpeasy2}
added.plot(ptsd~agility + speed + superpower + kills + north_south, data=avengers)
```

(These plots seem to say that, even after controlling for agility, speed, superpowers, and # of kills, which battlefield one fought on [north versus south] *still* predicts ptsd). 

Sometimes, however, you might want to have multiple interest variables. Or maybe you have quadratic and/or interaction effects you need to model for the control variables. In that case, you'll need to add the `lm.formula` argument to AVP:

```{r mvglmcomplex}
added.plot(ptsd~willpower + north_south, data=avengers, 
           lm_formula = ptsd~minutes.fighting + I(minutes.fighting^2))
```

The above example fits a quadratic (nonlinear) regression between `minutes.fighting` and `ptsd`, then plots willpower and north_south against these residuals. In other words, `minutes.fighting` (and the nonlinear effect of `minutes.fighting`) are the "control variables" and willpower and north_south are the interest variables. 

## Summary

For this book, "multivariate" means we have multiple independent variables. We use multivariate analyses when we want to model interactions, when we want to control for things, and/or when we want to predict things. If we're trying to visualize interactions, we can use flexplot to represent a variable as a color/symbol/line or as separate panels. If we're trying to visualize after we've controlled for variables, we use AVPs. 

## Practice

1. Describe how each of the following relationships will look. Try drawing a plot that matches your expectations. Then plot the actual relationships. How close were you?

```{r, eval=FALSE}
flexplot(superpower~1, data=avenger)
flexplot(speed~agility + superpower, data=avenger)
flexplot(agility~superpower | speed, data=avenger)
flexplot(agility~speed + strength | superpower, data=avenger)
flexplot(agility~speed | strength + superpower, data=avenger)
```

2. Suppose we wanted to plot the relationship between agility and speed, after controlling for superpower. How would you use an AVP to do that?

3. Suppose our interest variables are speed and agility, our outcome variable is ptsd, and our control variables are kills and injuries (and the interaction between the two). How would you use an AVP to visualize that?


